Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-11 write-ups/PLDI/paper.tex
Smoothing and surgery.

% lstlisting macros
\lstset{
  language=C, % choose the language of the code
  showspaces=false, % show spaces adding particular underscores
  showstringspaces=false, % underline spaces within strings
  showtabs=false, % show tabs within strings adding particular underscores
  tabsize=8, % sets default tabsize to 2 spaces
  captionpos=b, % sets the caption-position to bottom
  mathescape=true, % activates special behaviour of the dollar sign
  basicstyle=\scriptsize\tt,
  columns=fullfelxible,
  %commentstyle=\rmfamily\itshape,
  morekeywords={barrier,kernel,global,__device__,__syncthreads,__global__,bool,then},
  escapeinside={(*@}{@*)}
}

@@

\newcommand{\offerfork}{\mathsf{offer\_fork}}
\newcommand{\offerkill}{\mathsf{offer\_kill}}
\newcommand{\globalbarrier}{\mathsf{global\_barrier}}
\newcommand{\resizingglobalbarrier}{\mathsf{resizing\_global\_barrier}}
\newcommand{\getgroupid}{\mathsf{get\_group\_id}}
\newcommand{\getnumgroups}{\mathsf{get\_num\_groups}}
\newcommand{\getlocalid}{\mathsf{get\_local\_id}}
\newcommand{\getglobalid}{\mathsf{get\_global\_id}}
\newcommand{\getlocalsize}{\mathsf{get\_global\_id}}

@@

We use two examples of irregular data-parallel algorithms---a work
stealing queue and breadth-first graph search---to introduce the
industry-standard OpenCL programming model on which we base cooperative
kernels.

@@

@@

\begin{lstlisting}
kernel work_stealing(global Task *pools) {
  while (more_work(pools)) {
    int pool_id = get_group_id();
    Task *t = pop_or_steal(pools, pool_id);
    if (t) {
      // task processing may create more work
      process_task(t, pools, pool_id);
    }
  }
}
\end{lstlisting}

@@

  int level = 0;
  global nodes * in_nodes = n0;
  global nodes * out_nodes = n1;

@@

\paragraph{OpenCL kernels}

The threads executing a kernel are hierarchically organised.  A kernel
is executed by a number of \emph{workgroups}, and each workgroup
consists of a fixed number of threads.  Builtin functions
$\getlocalid()$ and $\getgroupid()$ return a thread's local id within
its workgroup and the id of the workgroup, respectively.\footnote{In general, an OpenCL kernel is organised as a multi-dimensional grid
of multi-dimensional workgroups, and the above functions each take
an integer dimension argument.  For simplicity, we consider only the
one-dimensional case throughout the paper, which captures all the
GPU-accelerated irregular algorithms we are aware of, and use
e.g.\ $\getglobalid()$ to mean $\getglobalid(0)$.
}  The number
of threads per workgroup and the number of workgroups are obtained via
$\getlocalsize()$ and $\getnumgroups()$.  A thread's local and global
id are related via the equation $\getglobalid() = \getgroupid() \times
\getlocalsize() + \getlocalid()$.

\ADComment{Maybe cut?} Within a workgroup, threads may also be
organised into \emph{subgroups} (termed \emph{warps} and
\emph{wavefronts} on NVIDIA and AMD architectures, respectively).

\paragraph{Memory spaces and memory model}
An OpenCL kernel has access to four memory spaces.  Shared virtual
memory (SVM) is accessible to all threads, and to the host
application.  Global memory is shared among all threads executing a
kernel, but can only be safely accessed by the host application before
or after kernel execution.  Local memory is shared among all threads
in the same workgroup: each workgroup has its own portion of local
memory, which can be used for fast intra-workgroup communication.
Each thread has an allocation of very fast private memory, where
function-local variables are allocated.

Fine-grained communication between threads, and also with the host in
the case of SVM, is enabled via a set of atomic data types and
operations, based on the C11 memory model~\cite{...}.  In particular,
atomic SVM operations enable fine-grained communication between host
and device.  Our implementation of cooperative kernels
(Section~\ref{sec:implementation}) depends on this recent OpenCL
feature.

\paragraph{Work stealing example}
Figure~\ref{fig:workstealing} shows a simplified version of an OpenCL
kernel that uses a work stealing queue.
%
The work stealing paradigm is useful when a computation can be split into multiple
tasks that can be processed independently, but where the number of
tasks is not known ahead of time because processing a given task may
create an arbitrary number of new tasks. Work stealing enables dynamic
balancing of tasks across several processing units, OpenCL workgroups
in our case. Each workgroup has an associated task pool from which it
obtains tasks to process, and to which it stores newly created
tasks. When a workgroup's pool becomes empty, a workgroup---rather
than terminating---tries to \emph{steal} a task from an other
pool. Hence, workgroups keep on processing tasks until all the work is
done.

\myfiglong\ref{fig:workstealing} illustrates an OpenCL implementation
of a work stealing algorithm. All threads start their execution at the
function marked with the $\mathsf{kernel}$ keyword. They receive a
pointer to the task pools, which reside in global memory, and which
were initialized by the host program beforehand to contain the initial
tasks. Each thread can identify the workgroup it belongs to via the
function $\getgroupid()$.  Here the group id is stored into the
variable $\mathsf{pool\_id}$ (which is, by default, in private memory)
and used to access the relevant task pool. The
$\mathsf{pop\_or\_steal}$ function is called by all threads of a
workgroup, although inside only the thread with a local id equal to
$0$, also called the \emph{master thread} of the workgroup,
\ADComment{Hugues, I changed; please check:} will try to pop a task
from the workgroup's pool, or steal a task from a different pool. The
master thread uses local memory inside $\mathsf{pop\_or\_steal}$ to
ensure that the returned value is valid for all threads of its
workgroup. If a task was obtained, then all threads of the workgroup
participate in the processing of the task inside the
$\mathsf{process\_task}$ function. \HEComment{Manage to say that a
  task is considered big enough to benefit from parallel computation}
\ADComment{I think that would be too much detail given the space we
  have, so I would not delve into this. Remove both comments if you
  agree.} Processing this task may lead to the creation of further
tasks, which the master thread will push to the workgroup's pool.

Although not depicted here, concurrent accesses to each pool inside
$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
mutex which is implemented using atomic compare and swap operations on
global memory.  As a result, the kernel presents two opportunities for
spin-waiting: spinning to obtain a mutex, and spinning in the main
kernel loop to obtain a task to process.  The algorithm thus depends
on a fair scheduler, which, as discussed in Section~\ref{sec:intro},
is not provided by current GPU programming models.  In
Section~\ref{sec:cooperativeexamples} we show how the example can be
updated to use our cooperative threads programming model to resolve
the scheduling issue.

\ADComment{Describe some
  aspects of the example, saying that the global barrier is not a
  primitive, and can be coded up using the memory model but is not
  safe due to the execution model.  Make it clear that the barrier
  uses busy-waiting.}

\paragraph{Graph traversal example} Figure~\ref{fig:graphsearch} shows the
skeleton of a frontier based graph traversal algorithm (this approach
has been shown to be efficient on GPUs in~\ref{...}). The kernel is
given three arguments in global memory: a graph structure, and two
arrays that can contain graph nodes. Initially, {\tt n0} contains the
starting nodes to process. Threads keep three local variables: {\tt
  level} which indicates the frontier level which is currently being
processed, {\tt in\_nodes} and {\tt out\_nodes} which point to
distinct node arrays. {\tt in\_nodes} contains the nodes to be
processed for the current frontier, while {\tt out\_nodes} contains
the nodes to be processed during the next frontier.

The application iterates as long as the current frontier contains
nodes to process (line~\ref{line:graph:iterate}). At each frontier,
the nodes to be processed are evenly distributed between the executing
threads. This is done through \emph{stride} based processing. A thread
retrieves its global thread id and the total number of threads
(obtained through the {\tt get\_global\_id()} and {\tt
  get\_global\_size()} OpenCL primitives). A thread then processes
each node associated with id (i.e.\ indices with its thread id value
added to any integer multiple of the total number of threads up to the
list size). A thread processes a node via the {\tt process\_node}
function, which may (a) push nodes to process in the next frontier to
the {\tt out\_nodes} list and (b) use the frontier level in
computation. After processing the frontier, the threads swap their
local node array pointers (line~\ref{line:graph:swap}), i.e.\ the {\tt
  out\_nodes} of the current frontier becomes the input of the next
frontier.

@@

The global barrier used in this application is not provided as a GPU
primitive, though many previous works have shown that such a global
barrier can be implemented~\cite{...}. These implementations use well
defined atomic memory operations to ensure memory consistency and use
spinning to ensure threads wait at the barrier for the remaining
threads to reach the barrier. Even though such barriers appear to run
reliably on current GPUs, they violate GPU language specifications,
which do not provide fair scheduling of workgroups. These barriers
require fair workgroup scheduling in order to function correctly. That
is, \emph{all} workgroups must be fairly scheduled in order to reach
the barrier, otherwise, the workgroups waiting at the barrier will not
make progress.

Two barrier primitives are provided in OpenCL.  A workgroup barrier
synchronises all threads in a workgroup and ensures memory
consistency.  This barrier is a \emph{workgroup-level} function: it
must be encountered uniformly by all or none of the threads of a
workgroup, thus can only appear in conditional code if all threads in
a workgroup that reach the barrier agree on the guards of all
enclosing conditional statements.  A subgroup barrier works
analogously at the subgroup level.  As discussed below, a
\emph{global} barrier (synchronising all threads running a kernel) is
not provided as a primitive.

HSA provides limited, one-way guarantees,
stating~\cite[p. 46]{HSAprogramming11}: \emph{Work-group A can wait
  for values written by work-group B without deadlock provided ... A
  comes after B in work-group flattened ID order}. However, this
one-way communication is not sufficient so support either the
work stealing or graph application examples of
Figures~\ref{fig:workstealing} and \ref{fig:graphsearch}. As described
above, both require \emph{symmetric} communication between
threads. \ADComment{Tyler, can you state the HSA guarantee?},
\TSComment{How is this?}~\cite{...}.

Revisiting the examples of Figures~\ref{fig:workstealing}
and~\ref{fig:graphsearch}: the mutex function of
Figure~\ref{fig:workstealing} will cause starvation if a workgroup
that holds the mutex is not scheduled again and so does not release
the mutex.  A similar issue affects the global barrier function of
Figure~\ref{fig:graphsearch}.  Because the barrier is implemented
using busy-waiting, a set of workgroups waiting at the barrier may be
continuously scheduled, starving a workgroup that has not yet arrived
at the barrier and meaning that no workgroup proceeds past the
barrier.

\section{The Cooperative Kernels Programming Model}\label{sec:cooperativekernels}

@@

As with a regular OpenCL kernel (see Section~\ref{sec:background}), a
cooperative kernel is launched by the host application.  The host
application passes parameters to the kernel and specifies a desired
number of workgroups, each consisting of a specified number of
threads.  The parameters to a cooperative kernel are \emph{immutable}, as if
they were qualified with the \texttt{const} keyword (though pointer
parameters can refer to mutable data).

@@

We now explain the semantics of these language extensions, after which
we explain how the work stealing and graph traveral algorithms of
Section~\ref{sec:background} can be turned into cooperative kernels
that use the extensions.  We present the semantics intuitively, using English.
In Appendix~\ref{appendix:semantics} we present an abstract operational semantics for our language extensions in a simple GPU-like programming model.

@@

The $\getgroupid()$ function retrieves the id of a workgroup as usual
(see Section~\ref{sec:background}).  When executed by a cooperative
kernel, $\getnumgroups()$ returns $M$, the \emph{current} number of
active workgroups.  This is in contrast to $\getnumgroups()$ for
regular kernels, which returns the fixed number of workgroups that
were requested at kernel launch time.

@@

Similar to the OpenCL workgroup $\mathsf{barrier}$ primitive,
$\offerkill$, is a workgroup-level function---it must be encountered
uniformly by all threads in a workgroup (see
Section~\ref{sec:background}).

@@

\ADComment{TODO.  We want to show the examples from Section~\ref{sec:background} ported to use cooperative kernels.  We want to make it clear that it is easy to place the primitives.}

>>>>

% lstlisting macros
\lstset{
  language=C, % choose the language of the code
  showspaces=false, % show spaces adding particular underscores
  showstringspaces=false, % underline spaces within strings
  showtabs=false, % show tabs within strings adding particular underscores
  tabsize=8, % sets default tabsize to 2 spaces
  captionpos=b, % sets the caption-position to bottom
  mathescape=true, % activates special behaviour of the dollar sign
  basicstyle=\scriptsize\tt,
  columns=fullfelxible,
  %commentstyle=\rmfamily\itshape,
  morekeywords={barrier,kernel,global,__device__,__syncthreads,__global__,bool,then},
  escapeinside={(*@}{@*)},
  numbers=left
}

@@

\newcommand{\offerfork}{\mathsf{offer\_fork}}
\newcommand{\offerkill}{\mathsf{offer\_kill}}
\newcommand{\globalbarrier}{\mathsf{global\_barrier}}
\newcommand{\resizingglobalbarrier}{\mathsf{resizing\_global\_barrier}}
\newcommand{\getgroupid}{\mathsf{get\_group\_id}}
\newcommand{\getnumgroups}{\mathsf{get\_num\_groups}}
\newcommand{\getlocalid}{\mathsf{get\_local\_id}}
\newcommand{\getglobalid}{\mathsf{get\_global\_id}}
\newcommand{\getlocalsize}{\mathsf{get\_global\_id}}
\newcommand{\getglobalsize}{\mathsf{get\_global\_size}}

\newcommand{\keyword}[1]{\mathsf{#1}}

@@

We provide background on the industry-standard OpenCL programming
model on which we base cooperative kernels (\mysec\ref{sec:opencl}),
and illustrate OpenCL using two irregular data-parallel algorithms---a
work stealing queue and breadth-first graph search---that also
pinpoint the problems associated with unfair scheduling of workgroups
and motivate our cooperative threads approach (\mysec\ref{sec:openclexamples}.

\subsection{OpenCL Background}

@@

The threads executing a kernel are hierarchically organised.  A kernel
is executed by a number of \emph{workgroups}, and each workgroup
consists of a fixed number of threads.  Builtin functions
$\getlocalid()$ and $\getgroupid()$ return a thread's local id within
its workgroup and the id of the workgroup, respectively.\footnote{In general, an OpenCL kernel is organised as a multi-dimensional grid
of multi-dimensional workgroups, and the above functions each take
an integer dimension argument.  For simplicity, we consider only the
one-dimensional case throughout the paper, which captures all the
GPU-accelerated irregular algorithms we are aware of, and use
e.g.\ $\getglobalid()$ to mean $\getglobalid(0)$.
}  The number
of threads per workgroup and the number of workgroups are obtained via
$\getlocalsize()$ and $\getnumgroups()$.  A thread's local and global
id are related via the equation $\getglobalid() = \getgroupid() \times
\getlocalsize() + \getlocalid()$.  The total number of threads executing the kernel is given by $\getglobalsize()$, which is equal to $\getlocalsize()\times\getnumgroups()$.

The execution of all threads in a workgroup can be synchronised via a
workgroup barrier, which also ensures memory consistency.  This
barrier is a \emph{workgroup-level} function: it must be encountered
uniformly by all or none of the threads of a workgroup, thus can only
appear in conditional code if all threads in a workgroup that reach
the barrier agree on the guards of all enclosing conditional
statements.  As discussed below, a \emph{global} barrier
(synchronising all threads running a kernel) is \emph{not} provided as
a primitive.

\paragraph{Memory spaces and memory model}
An OpenCL kernel has access to four memory spaces.  \emph{Shared virtual
memory} (SVM) is accessible to all threads and the host application concurrently.  \emph{Global} memory is shared among all threads executing a
kernel, but can only be safely accessed by the host application before
or after kernel execution.  Each workgroup has its own portion of \emph{local},
which can be used for fast intra-workgroup communication, and every thread has
an allocation of very fast \emph{private} memory, where
function-local variables are allocated.

Fine-grained communication between threads, and also with the host in
the case of SVM, is enabled via a set of atomic data types and
operations, based on the C11 memory model~\cite{...}.  \ADComment{Could cut:} In particular,
atomic SVM operations enable fine-grained communication between host
and device.  Our implementation of cooperative kernels
(Section~\ref{sec:implementation}) depends on this recent OpenCL
feature.

HSA provides limited, one-way guarantees,
stating~\cite[p. 46]{HSAprogramming11}: \emph{Work-group A can wait
  for values written by work-group B without deadlock provided ... A
  comes after B in work-group flattened ID order}. However, this
one-way communication is not sufficient so support either the
work stealing or graph application examples of
Figures~\ref{fig:workstealing} and \ref{fig:graphsearch}. As described
above, both require \emph{symmetric} communication between
threads.

\subsection{Motivating Examples}\label{sec:openclexamples}

@@

\begin{lstlisting}
kernel work_stealing(global Task * pools) {
  while (more_work(pools)) {
    int pool_id = get_group_id();
    Task * t = pop_or_steal(pools, pool_id);
    if (t) {
      // task processing may create more work
      process_task(t, pools, pool_id);
    }
  }
}
\end{lstlisting}

@@

\begin{lstlisting}
kernel graph_app(global graph * g, 
                 global nodes * n0, global nodes * n1) {
  int level = 0;
  global nodes * in_nodes = n0;
  global nodes * out_nodes = n1;

@@

\paragraph{Work stealing example}
%
The work stealing paradigm is useful when a computation can be split into multiple
tasks that can be processed independently, but where the number of
tasks is not known ahead of time because processing a given task may
create an arbitrary number of new tasks. Work stealing enables dynamic
balancing of tasks across several processing units, OpenCL workgroups
in our case. Each workgroup has an associated task pool from which it
obtains tasks to process, and to which it stores newly created
tasks. When a workgroup's pool becomes empty, a workgroup---rather
than terminating---tries to \emph{steal} a task from an other
pool. Workgroups keep on processing tasks until all the work is
done.

\myfiglong\ref{fig:workstealing} shows a simplified version of an OpenCL
kernel that uses a work stealing queue. All threads start their execution at the
function marked with the $\mathsf{kernel}$ keyword. They receive a
pointer to the task pools, which reside in global memory, and which
were initialized by the host program beforehand to contain the initial
tasks. Each thread can identify the workgroup it belongs to via the
function $\getgroupid()$.  Here the group id is stored into the
variable $\mathsf{pool\_id}$ (which is, by default, in private memory)
and used to access the relevant task pool. The
$\mathsf{pop\_or\_steal}$ function is called by all threads of a
workgroup.  Inside $\mathsf{pop\_or\_steal}$ only the thread with a local id equal to
$0$, also called the \emph{master thread} of the workgroup,
\ADComment{Hugues, I changed; please check:} will try to pop a task
from the workgroup's pool, or steal a task from a different pool. The
master thread uses local memory inside $\mathsf{pop\_or\_steal}$ to
ensure that the returned value is valid for all threads of its
workgroup. If a task was obtained, then all threads of the workgroup
participate in the processing of the task via the
$\mathsf{process\_task}$ function. \HEComment{Manage to say that a
  task is considered big enough to benefit from parallel computation}
\ADComment{I think that would be too much detail given the space we
  have, so I would not delve into this. Remove both comments if you
  agree.} Processing this task may lead to the creation of further
tasks, which the master thread will push to the workgroup's pool.

Although not depicted here, concurrent accesses to each pool inside
$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
mutex, implemented using atomic compare and swap operations on
global memory.  As a result, the kernel presents two opportunities for
spin-waiting: spinning to obtain a mutex, and spinning in the main
kernel loop to obtain a task to process.  The algorithm thus depends
on a fair scheduler, which, as discussed in Section~\ref{sec:intro},
is not provided by current GPU programming models.

\paragraph{Graph traversal example} Figure~\ref{fig:graphsearch} shows the
skeleton of a frontier-based graph traversal algorithm; such algorithms have
been shown to be execute efficiently on GPUs~\cite{...}.
The kernel is
given three arguments in global memory: a graph structure, and two
arrays that can contain graph nodes. Initially, $\keyword{n0}$ contains the
starting nodes to process. Threads keep three local variables: $\keyword{level}$, which indicates the frontier level currently being
processed, and $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$, which point to
distinct node arrays. The nodes to be
processed for the current frontier are contained in $\keyword{in\_nodes}$, while $\keyword{out\_nodes}$ contains
the nodes to be processed during the next frontier.

The application iterates as long as the current frontier contains
nodes to process (line~\ref{line:graph:iterate}). At each frontier,
the nodes to be processed are evenly distributed between the executing
threads. This is done through \emph{stride} based processing. A thread
retrieves its global thread id and the total number of threads
(obtained through the $\getglobalid()$ and $\getglobalsize()$ primitives). A thread then processes
each node associated with its id, i.e.\ nodes at indices that are a multiple of the total number of threads, plus the thread's id. A thread processes a node via the $\keyword{process\_node}$
function, which may (a) push nodes to process in the next frontier to
$\keyword{out\_nodes}$ and (b) use the current frontier, $\keyword{level}$, in
the computation. \ADComment{Tyler: should swap take the addresses of the pointers?} After processing the frontier, the threads swap their
local node array pointers (line~\ref{line:graph:swap}), i.e.\ the $\keyword{out\_nodes}$ of the current frontier becomes the $\keyword{in\_nodes}$ of the next
frontier.

@@

The global barrier used in this application is not provided as a GPU
primitive, though many previous works have shown that such a global
barrier can be implemented~\cite{...}. These implementations use well
defined atomic memory operations to ensure memory consistency, and
employ spinning to ensure threads wait at the barrier until all
threads have arrived.  This spinning behaviour requires fair
scheduling between workgroups for the barrier to operate correctly,
and as a result the barrier is \emph{not} guaranteed to work correctly
under current GPU programming models.  Such barriers appear to run
reliably on current GPUs for kernels that are executed with no more
workgroups than there are compute units; this is due to the fairness
of the occupancy-bound execution model that current GPUs appear to
provide.  But, as discussed in \mysec\ref{sec:intro}, this
occupancy-bound model is not mandated by language standards or vendor
implementations, and is unlikely to be guaranteed in the future.

In Section~\ref{sec:cooperativeexamples} we show how the work stealing
and graph traversal examples of Figures~\ref{fig:workstealing} and~\ref{fig:graphsearch} can be
updated to use our cooperative kernels programming model to resolve
the scheduling issue.

\section{Cooperative Kernels}\label{sec:cooperativekernels}

@@

As with a regular OpenCL kernel (see Section~\ref{sec:opencl}), a
cooperative kernel is launched by the host application.  The host
application passes parameters to the kernel and specifies a desired
number of workgroups, each consisting of a specified number of
threads.  The parameters to a cooperative kernel are \emph{immutable}, as if
they were qualified with the \texttt{const} keyword (though pointer
parameters can refer to mutable data).

@@

We now explain the semantics of these language extensions, after which
we explain how the work stealing and graph traveral algorithms of
Section~\ref{sec:openclexamples} can be turned into cooperative kernels
that use the extensions.  We present the semantics intuitively, using English.
In Appendix~\ref{appendix:semantics} we present an abstract operational semantics for our language extensions in a simple GPU-like programming model.

@@

The $\getgroupid()$ function retrieves the id of a workgroup as usual
(see Section~\ref{sec:opencl}).  When executed by a cooperative
kernel, $\getnumgroups()$ returns $M$, the \emph{current} number of
active workgroups.  This is in contrast to $\getnumgroups()$ for
regular kernels, which returns the fixed number of workgroups that
were requested at kernel launch time.

@@

Similar to the OpenCL workgroup $\mathsf{barrier}$ primitive,
$\offerkill$, is a workgroup-level function---it must be encountered
uniformly by all threads in a workgroup (see
Section~\ref{sec:opencl}).

@@

\ADComment{TODO.  We want to show the examples from Section~\ref{sec:openclexamples} ported to use cooperative kernels.  We want to make it clear that it is easy to place the primitives.}
